What can GitHub and Googlecode Issues tell us about Exceptions in Android and Java?

### 2014-08-31 android-stacks.tex
changes all over

\subsection{What the Common Root Exceptions can tell us? }

@@

\subsubsection{The high number of NullPointerExceptions}
From this set the java.lang.NullPointerException was the most reported root cause; 1225 stack traces 
had a java.lang.NullPointerException  as root cause, which corresponds of 30,08\%  of 
all analyzed stack traces. Considering the number of distinct project repositories, 
the java.lang.NullPointerException, was reported on at least one issue of 99 projects 
(which represents 20,54\% of all projects reporting at least one stack trace).

 The NullPointerExceptions are mainly signaled inside Android platform and inside 
application code, although we also find
NullPointerExceptions being signaled from third-party libraries. Regarding
reusable code, there is no consensus whether it is a good or bad practice to 
re-throw a NullPointerException. Some prefer to encapsulate such an exception on
IllegalArgumentException, while others~\cite{bloch2008effective} argue that the
NullPointerException makes the cause of the problem explicit and hence 
should not be wrapped.

The high prevalence of NullPointerExceptions is aligned with the findings other 
works~\cite{kim2013predicting,fraser20131600,csallner2004jcrasher}. For instance, Sunghun et
al.~\cite{kim2013predicting} showed that in Eclipse bug report system 38\% the bugs 
related to exception handling were caused by NullPointerException; other works on robustness 
testing~\cite{maji2012empirical,csallner2004jcrasher} showed that most of the automatically 
detected bugs were due to NullPointerExceptions and implicitly-signaled of Java
environment (as the ones found in this study).

In Section XX we provide further discussion on the NullPointerException problem, pointing 
to ways of how to deal with it and mitigate it.

\subsubsection{Direct instances of RuntimeExceptions being thrown.}

From Tables~\ref{tab:toptenjava} and~\ref{tab:toptenandroid} we can observe that,
in both Java and Android related issues, direct instances
java.lang.RuntimeException were thrown  in 10,94\% of Java repositories and
18,88\% of Android repositories respectively. 
In the Android repositories, most of such exceptions  were  thrown by the Android 
platform/OS (167 out of 232).

@@

\subsubsection{Classifying Top 100 Exceptions}

To get a broader view of the root causes of stack traces, we performed an empirical evaluation
whose goal was to categorized the top 100 root exceptions. The top 100 root causes correspond 
to approximately 95\% of all exception types reported on the exception stack-issues.

@@

\begin{table}
  \centering
  \begin{tabular}{lrr}
    \hline
    \bfseries{Category} &  \bfseries{\% reported on stacks} \\
    \hline
      Programming logic (java.lang and util) &  57,9\%\\ %2326
      Resources (IO)                         &   18,1\% \\ %727
      Security                               &  4,1\%\\ %165 
      Concurrency                            &  2,9\% \\ %116 
      Backward compatibility                 & 5,5\% \\ %219
      Specific Exceptions               &  4,9\%\\ %197 
      General (Error, Exception, Runtime)    &  6,7\%\\ %267 
    \hline
  \end{tabular}
  \caption{Characterization of the 100 root causes.}
  \label{tab:tophundrend}
\end{table}

@@

In this second analysis the programming mistakes were also the
most mentioned root causes, followed by exceptions related to resource 
management (which was expected since on mobile environment 
resources are usually scarce). Some of such programming mistakes,
however, may be interactly related to one of the other concerns. For 
instance, a concurrency problem may lead to a NullPointerException.
This classification however, was only based on the semantic embedded 
on the exception types.

@@

\subsection{What the Types of Root Exceptions can tell us?}
%\subsection{Exploring the Information Embedded on Exception Types}
%\subsection{The Exception Types and What They can Tell us? }

@@

We could also observe checked exceptions (which according to Java specification represent
recoverable conditions) as the root causes of reported exception stack traces.
A checked exception can remain uncaught if it is declared by every method on the call chain or if 
it is wrapped in a runtime exception and re-thrown. Next, section describes the kinds of 
wrappings found in this study.

\subsection{What the Exception Wrappings can tell us?}
%\subsection{Exploring the Information Embedded on Exception Wrappings}
Table~\ref{tab:wrappingandroid} presents the wrappings found in this study for all
stack traces found in Android repositories. As we can see, some of the checked
exceptions were indeed wrapped in runtime exceptions or even errors.

@@

\subsubsection{Runtime wrapping Checked Exception}

Such wrapping was detected on 70 distinct projects. Inspecting the 
elements responsible for performing such wrapping, we could observe that
most of such wrappings were performed on methods defined on Android
Platform (approximately 58%).

@@

\subsubsection{Runtime Exception wrapping Error}

We could find  57 stack traces on 38 projects on which an Error was wrapped in a runtime exception.
Manually inspecting each exception stack trace, we could observe that 44% of the wrappings were performed 
by android.os.AsyncTask. This class is responsible for dealing with asynchronous tasks in Android.
The code snippet below shows that any throwable signaled in the context of an asynchronous 
task is converted in to a runtime and re-thrown.

@@

  \begin{tabular}{|p{7.9cm} p{0.1cm}|}
%\begin{tabular}{ll}
    \hline
 \bfseries{Runtime-Error Wrapping Examples} &   \\  %\bfseries{\#}
    \hline
java.lang.RuntimeException - java.lang.OutOfMemoryError & \\ %30
java.lang.RuntimeException - java.lang.NoClassDefFoundError &\\ %6
java.lang.RuntimeException - java.lang.UnsatisfiedLinkError) & \\ %5
java.lang.RuntimeException -  java.lang.StackOverflowError)  & \\ %4
\hline
\bfseries{Checked-Runtime Wrapping Examples} &  \\
 \hline
...android.robospice.CacheSavingException - java.lang.ClassCastException &  \\ %6
java.lang.reflect.InvocationTargetException - java.lang.NullPointerException &   \\ %3
java.lang.reflect.InvocationTargetException - ...ResourcesdNotFoundException &  \\ %2
java.sql.SQLException - android.database.sqlite.SQLiteDiskIOException &  \\ %2
 \hline
\bfseries{Checked-Error Wrapping Examples} & \\
 \hline
java.lang.reflect.InvocationTargetException - java.lang.OutOfMemoryError &  \\ %5
java.lang.Exception - java.lang.OutOfMemoryError &  \\ %1
java.lang.reflect.InvocationTargetException - java.lang.NoSuchMethodError &	 \\ %1
java.lang.reflect.InvocationTargetException - java.lang.UnsatisfiedLinkError&	 \\ %1
\hline
\bfseries{Error-Checked Wrapping Examples} & \\
 \hline
java.lang.NoClassDefFoundError - java.lang.ClassNotFoundException &  \\ %4
java.lang.AssertionError - javax.crypto.ShortBufferException) &  \\ %3
\hline
\bfseries{Error-Runtime Wrapping Examples} &   \\
 \hline
java.lang.ExceptionInInitializerError - java.lang.NullPointerException &  \\ %9
java.lang.ExceptionInInitializerError - java.lang.IllegalArgumentException &	 \\ %2
 \hline
  \end{tabular}
\caption{Wrapping examples}
\label{tab:wrapping01}
\end{table}

\subsubsection{Checked wrapping Runtime Exception}

Manually inspecting the stack traces we could observe that most of the wrappings were performed 
by RoboSpice (an android library that supports the creation of asynchronous long running tasks.
This library has 1.682 stars and 367 forks) and by the reflection Java library (used by Android, 
Application and library classes). Inspecting the code related to such wrappings we could observe that
the reflection library convert every exception that happens 
when invoking a reflected method into an InvoncationTargetException (this conversion is 
performed inside native code).

@@

\subsubsection{Checked Exception wrapping Error}

Almost all of these wrappings were also caused by the reflection library 
used by Android applications. The method responsible for the wrappings
were native methods written in C. Table  ~\ref{tab:wrapcheckerr} illustrates
some oth these wrappings some of them are masking an OutOfMemoryError
into a checked exception.

@@

\subsubsection{Error wrapping Checked Exception}

Some of these wrappings were performed by methods defined by java.lang.Class methods such as 
at java.lang.Class.getDeclaredFields(Native Method)
at java.lang.Class.getDeclaredConstructors(Native Method), 
while others were performed by javax.crypto package. Table X illustrates
the exception types involved in such wrappings.

@@

\subsubsection{Error wrapping Runtime}

After manually inspecting the code associated to such wrappings we discovered that all of them were 
caused by static initializers. If any exception is thrown in the context of a static initializer (i.e., static block) 
it is converted into an ExceptionInitializerError on the point where the class is first used. Table X illustrates
the exception types involved in such wrappings.

@@

\subsubsection{Multiple Wrappings}

We could also observe that in 205 stacks found in Android repository the exception was wrapped 
more then once. Figure ~\ref{fig:sizewrapbhists}  illustrates the number of wrappings per stack, both
 in Java and Android studied repositories. One interesting multiple-wrapping found in Android repo was the 
following: Runtime-Checked-Runtime-Checked-Runtime-Checked-Runtime.

@@

\emph{"Everybody hates thinking about exceptions,                   							because they’re not supposed to happen."}
\textbf{Brian Foote}

@@

\subsection{On the need of a rescue to the almost Infeasible task of preventing uncaught exceptions}
In this study we could observe undocumented runtime exceptions thrown by third party code,
and even undocumented checked exception thrown by Java supporting library.
Such undocumented exceptions make difficult, and most of the times infeasible
for the client code to protect against "unforeseen" situations that may happen 
while calling a library code.

\subsection{On the Null Pointer Problem}

@@

"I call it my billion-dollar mistake. It was the invention of the null reference in 1965 [...] I couldn’t resist the temptation to put 
in a null reference, simply because it was so easy to implement. This has led to innumerable errors, vulnerabilities, and system 
crashes, which have probably caused a billion dollars of pain and damage in the last forty year."

>>>>

\subsection{What can the Common Root Exceptions tell us? }

@@

\emph{\textbf{The high number of NullPointerExceptions.}}
From this set the java.lang.NullPointerException was the most reported root cause; 1225 stack traces 
had a java.lang.NullPointerException  as root cause, which corresponds of 30,08\%  of 
all analyzed stack traces. Considering the number of distinct project repositories, 
the java.lang.NullPointerException, was reported on at least one issue of 99 projects 
(which represents 20,54\% of all projects reporting at least one stack trace).
 The NullPointerExceptions are mainly signaled inside Android platform and inside 
application code, although we also find
NullPointerExceptions being signaled from third-party libraries. Regarding
reusable code, there is no consensus whether it is a good or bad practice to 
re-throw a NullPointerException. Some prefer to encapsulate such an exception on
IllegalArgumentException, while others~\cite{bloch2008effective} argue that the
NullPointerException makes the cause of the problem explicit and hence 
should not be wrapped.

The high prevalence of NullPointerExceptions is aligned with the findings other 
works~\cite{kim2013predicting,fraser20131600,csallner2004jcrasher}. For instance, Sunghun et
al.~\cite{kim2013predicting} showed that in Eclipse bug report system 38\% the bugs 
related to exception handling were caused by NullPointerException; other works on robustness 
testing~\cite{maji2012empirical,csallner2004jcrasher} showed that most of the automatically 
detected bugs were due to NullPointerExceptions and implicitly-signaled of Java
environment (as the ones found in this study).In Section XX we provide further 
discussion on the NullPointerException problem, pointing 
to ways of how to mitigate it.

\emph{\textbf{Direct instances of RuntimeExceptions being thrown.}}
From Table ~\ref{tab:toptenandroid} we can observe that direct instances
java.lang.RuntimeException were thrown in 18,88\% of Android repositories, and
most of such exceptions  were  thrown by the Android platform (167 out of 232).

@@

\emph{\textbf{Classifying Top 100 Exceptions.}} To get a broader view of the root causes of stack traces,
 we performed an empirical evaluation
whose goal was to categorized the top 100 root exceptions. The top 100 root causes correspond 
to approximately 95\% of all exception types reported on the exception stack-issues.

@@

\begin{table}
  \centering
  \begin{tabular}{lrr}
    \hline
    \bfseries{Category} &  \bfseries{\% reported on stacks} \\
    \hline
      Programming logic (java.lang and util) &  52,0\%\\ 
      Resources (IO, Memory, Batery)       &   23,9\% \\ 
      Security                               &  4,1\%\\  
      Concurrency                            &  2,9\% \\ 
      Backward compatibility                 & 5,5\% \\ 
      Specific Exceptions               &  4,9\%\\ 
      General (Error, Exception, Runtime)    &  6,7\%\\
  \end{tabular}
  \caption{Characterization of the 100 root causes.}
  \label{tab:tophundrend}
\end{table}

@@

Analizing the Top 100 exceptions the programming mistakes
were also the most reported. However, they may be interactly related 
to one of the other concerns. For  instance, a concurrency problem may 
lead to a NullPointerException. This classification however, 
was only based on the semantic embedded  on the exception types.

\noindent \fbox{ \begin{minipage}{0.96\columnwidth} \emph{
Programming mistakes were reported as the root cause of  52\% of the analyzed stack traces. 
In 30,08\% NullPointerExceptions was reported as the root cause.
}
\end{minipage}}

@@

\subsection{What can the Types of Root Exceptions tell us?}
%\subsection{Exploring the Information Embedded on Exception Types}
%\subsection{The Exception Types and What They can Tell us? }

@@

We could also observe checked exceptions (which according to Java specification represent
recoverable conditions) as the root causes of reported exception stack traces.
Two possible ways of a checked exception can remain uncaught are: the checked 
exception is declared by every method on the call chain or if it is wrapped in a 
runtime exception and re-thrown. However the inspection of stack traces 
and the signatured of methods revealed the a third option: an undocumented
checked exception thrown by a native method.

The code snippet in Figure X illustrate this scenario.
 The java/lang/Class declares a native method named getDeclaredMethods. 
The Java-side declaration of of this method does not have any throws clause, 
leading programmers and the compiler to think that no checked exceptions can be thrown.

 However, the C-code implementation did throw a "checked exception" NoSuchMethodException, 
violating the declaration. The Java compiler, could not detect this violation because does 
not perform static exception  checking on native methods. This type of bugs is hard to diagnose
because the developer usualy do not have access to the native code (written in other language).

Consequently, since it was not expected by the programmer, when such method throws 
this exception, such undocumented exception may remain
uncaught and cause the app to crash, or be mistakenly handled by subsumption - 
The exception stack traces reporting this scenario is actually a real bug of Android 
Gingerbread version (which still accounts for 13.6% of devices running Android).

\noindent \fbox{ \begin{minipage}{0.96\columnwidth} \emph{
  0,8\% of the methods documented the explicitly signaled runtime exceptions.
  The study also detected an undocumented checked exception signaled by
  a native method defined in Android.
}
\end{minipage}}

\subsection{What can the Exception Wrappings tell us?}
%\subsection{Exploring the Information Embedded on Exception Wrappings}
Table~\ref{tab:wrappingandroid} presents the wrappings found in this study for all
stack traces found in Android repositories. As we can see, some of the checked
exceptions were indeed wrapped in runtime exceptions or even errors.

@@

\emph{\textbf{Runtime wrapping Checked Exception.}} Such wrapping was 
detected on 70 distinct projects. Inspecting the 
elements responsible for performing such wrapping, we could observe that
most of such wrappings were performed on methods defined on Android
Platform (approximately 58%).

@@

\emph{\textbf{Runtime Exception wrapping Error.}} We could find  57 stack traces 
on 38 projects on which an Error was wrapped in a runtime exception.
Manually inspecting each exception stack trace, we could observe that 44% of the wrappings were performed 
by android.os.AsyncTask. This class is responsible for dealing with asynchronous tasks in Android.
The code snippet below shows that any throwable signaled in the context of an asynchronous 
task is converted in to a runtime and re-thrown.

@@

 %\begin{tabular}{|p{7.9cm} p{0.1cm}|}
\begin{tabular}{ll}
    \hline
 \bfseries{Runtime-Error Wrapping Examples} &   \\  %\bfseries{\#}
    \hline
java.lang.RuntimeException - java.lang.OutOfMemoryError & \\ %30
java.lang.RuntimeException - java.lang.NoClassDefFoundError &\\ %6
java.lang.RuntimeException - java.lang.UnsatisfiedLinkError) & \\ %5
java.lang.RuntimeException -  java.lang.StackOverflowError)  & \\ %4
\hline
\bfseries{Checked-Runtime Wrapping Examples} &  \\
 \hline
...android.robospice.CacheSavingException - java.lang.ClassCastException &  \\ %6
java.lang.reflect.InvocationTargetException - java.lang.NullPointerException &   \\ %3
java.lang.reflect.InvocationTargetException - ...ResourcesdNotFoundException &  \\ %2
java.sql.SQLException - android.database.sqlite.SQLiteDiskIOException &  \\ %2
 \hline
\bfseries{Checked-Error Wrapping Examples} & \\
 \hline
java.lang.reflect.InvocationTargetException - java.lang.OutOfMemoryError &  \\ %5
java.lang.Exception - java.lang.OutOfMemoryError &  \\ %1
java.lang.reflect.InvocationTargetException - java.lang.NoSuchMethodError &	 \\ %1
java.lang.reflect.InvocationTargetException - java.lang.UnsatisfiedLinkError&	 \\ %1
\hline
\bfseries{Error-Checked Wrapping Examples} & \\
 \hline
java.lang.NoClassDefFoundError - java.lang.ClassNotFoundException &  \\ %4
java.lang.AssertionError - javax.crypto.ShortBufferException) &  \\ %3
\hline
\bfseries{Error-Runtime Wrapping Examples} &   \\
 \hline
java.lang.ExceptionInInitializerError - java.lang.NullPointerException &  \\ %9
java.lang.ExceptionInInitializerError - java.lang.IllegalArgumentException &	 \\ %2
 \hline
  \end{tabular}
\caption{Examples of Cross-type wrappings}
\label{tab:wrapping01}
\end{table}

\emph{\textbf{Checked wrapping Runtime Exception.}} Manually inspecting the stack 
traces we could observe that most of the wrappings were performed 
by RoboSpice (an android library that supports the creation of asynchronous long running tasks.
This library has 1.682 stars and 367 forks) and by the reflection Java library (used by Android, 
Application and library classes). Inspecting the code related to such wrappings we could observe that
the reflection library convert every exception that happens 
when invoking a reflected method into an InvoncationTargetException (this conversion is 
performed inside native code).

@@

\emph{\textbf{Checked Exception wrapping Error.}} Almost all of these wrappings
 were also caused by the reflection library 
used by Android applications. The method responsible for the wrappings
were native methods written in C. Table  ~\ref{tab:wrapcheckerr} illustrates
some oth these wrappings some of them are masking an OutOfMemoryError
into a checked exception.

@@

\emph{\textbf{Error wrapping Checked Exception.}} Some of these wrappings were performed by methods defined by java.lang.Class methods such as 
at java.lang.Class.getDeclaredFields(Native Method)
at java.lang.Class.getDeclaredConstructors(Native Method), 
while others were performed by javax.crypto package. Table X illustrates
the exception types involved in such wrappings.

@@

\emph{\textbf{Error wrapping Runtime.}} After manually inspecting the code associated to such wrappings we discovered that all of them were 
caused by static initializers. If any exception is thrown in the context of a static initializer (i.e., static block) 
it is converted into an ExceptionInitializerError on the point where the class is first used. Table X illustrates
the exception types involved in such wrappings.

@@

\emph{\textbf{Multiple Wrappings.}} We could also observe that in 205 stacks found
 in Android repository the exception was wrapped 
more then once. One interesting multiple-wrapping found in Android repo was the 
following: Runtime-Checked-Runtime-Checked-Runtime-Checked-Runtime.

\noindent \fbox{ \begin{minipage}{0.96\columnwidth} \emph{
Odd wrappings were detected (e.g., a checked exception wrapping an OutOfMemoryError)
reveal fragilities on Java Exception Model.
}
\end{minipage}}

@@

\emph{``Everybody hates thinking about exceptions, because they are not supposed to happen."}
\textbf{Brian Foote}

\subsection{On the need of a rescue to the almost infeasible task of preventing uncaught exceptions}
In this study we could observe undocumented runtime exceptions thrown by third party code,
and even undocumented checked exception thrown by Java supporting library.
Such undocumented exceptions make difficult, and most of the times infeasible
for the client code to protect against `unforeseen" situations that may happen 
while calling a library code.

@@

\subsection{On the null pointer problem}

@@

\emph{``I call it my billion-dollar mistake. It was the invention of the null reference in 1965 [...] I couldn’t resist the temptation to put 
in a null reference, simply because it was so easy to implement. This has led to innumerable errors, vulnerabilities, and system 
crashes, which have probably caused a billion dollars of pain and damage in the last forty year."}
